DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems

### 2012-11-26 Paper_May_2012/main.tex
Accept with last minor todo: in Christian's Drag Formal ist meiner Ansicht nach ein Fehler. Ich habe einen Vorschlag als Todo mit Literaturangabe. Bitte anschauen. Ansonsten sind wir nun fertig bis auf Rechtschreibpruefung

\subsubsection{\texttt{localfunc.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
\todo{Habe ich mal umgeschrieben, bitte nochmal ansehen~WW.}
The main difference, and common to all functionals is that we only 
want to evaluate a functional, i.e., calculate a single number, and not a
residual-vector\todocg{, i.e., calculate the elements of a vector Diesen Teil wuerde ich streichen.}. Thus, we 
just loop over the quadrature points and not over the degrees of freedom 
on the face. In the present situation, we want to evaluate a functional 
that consists of a boundary integral. Thus we need to evaluate 
integrals on faces, that lie on $\Gamma_{circ}$, instead of integrals over
elements. 
This face contributions are calculated in \texttt{BoundaryValue} and added to 
sum up to the value of $J$.

@@

Now we may discretize the problem with some time-stepping scheme, which will 
remove the temporal derivative and replace it by some difference quotients. 
Since the exact way of doing this depends on the time-stepping scheme 
we need a new object that takes the user-given description of the PDE and 
generates the appropriate combinations of spatial integrals and scaling.
Thus a generic problem may look as follows:
\begin{lstlisting}
  template<typename PROBLEM>
      class TimeStepProblem(Problem & P);
\end{lstlisting}
At present we have prepared the following time-stepping schemes in \dope:
\begin{itemize}
\item Forward-(or Explicit)-Euler
\item Backward-(or Implicit)-Euler
\item Crank-Nicolson scheme
\item shifted Crank-Nicolson scheme (to deal with singular initial data)
\item fractional-step-$\theta$ scheme
\end{itemize}
\todo{bitte gegenlesen}
After choosing the time-stepping scheme, the SpaceTimeHandler  
loops over all time-steps, where in each of the 
time-steps we can reuse the functionality we have provided to solve 
stationary PDEs. 
This means for instance, if we have implemented the 
cell-integrals in the weak formulation for the stationary case, 
then we can use the exact same 
implementation again for the solution of the nonstationary version of 
the equation. However, we will possibly need to provide an additional 
term to cope with the integral for the time derivative.

@@

\subsection{Goal-oriented mesh refinement for Navier-Stokes}
In this example we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in Fig.~\ref{fig:example_ns} and the equation is discretized with the Taylor-Hood  element. We want to compute the drag force on the enclosed cylinder
\begin{equation}
J(u) = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nv _1 - pn,
\end{equation}
with the solution $u = (v,p)$, the normal $n$ and the viscosity $\nu = 0.001$.
\begin{figure}[hbt]
\centering
{\includegraphics[width=0.9\textwidth]{Pictures/local_grid_NS.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}
To achieve an efficient computation of the quantity $J$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}. The method delivers error indicators and by this way computational grids tailored to minimize the error in the previously given functional $J$. The price we have to pay is the additional computation of a (linear) dual problem, which is implemented as described in section~\ref{sec:opt}. In Fig.~\ref{fig:NS_local_grid} we see a resulting triangulation after eight refinement steps with roughly $320\,000$ degrees of freedom.

@@

\section{Conclusions}
\todo{Frage an CG und Winni: Guter Abschluss?}
In this article, we described the architecture 
of the novel software library DOpElib for solving 
PDEs and optimization problems.  The key idea behind 
the algorithms is the computation of some target 
functional \todocg{Ich wuerde eventuell die userfreundlichkeit mehr in den Fokus ruecken? The key idea of the software is to supply a toolkit which allows for fast implementation of the problems at hand, but at the same time gives the user full flexibility to change or alter the used algorithms}\todo{Bin daf\"ur~WW}. This is achieved by a unified interface 
to state of the art time-stepping methods, nonlinear solvers and optimization
routines. The capabilities of DOpElib are demonstrated with the help of three examples.

>>>>

\subsubsection{\texttt{localfunc.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
The main difference, and common to all functionals is that we only want to
\todocg{ evaluate a functional, i.e., <-wuerde ich streichen} calculate a single number, and not a
whole 
residual-vector. Thus, we 
just loop over the quadrature points and not over the degrees of freedom 
on the face. In the present situation, we want to evaluate a functional 
that consists of a boundary integral. Thus we need to evaluate 
integrals on faces, that lie on $\Gamma_{circ}$, instead of integrals over
elements. 
This face contributions are calculated in \texttt{BoundaryValue} and added to 
sum up to the value of $J$.

@@

Now we may discretize the problem with some time-stepping scheme, which will 
remove the temporal derivative and replace it by some difference quotients. 
Since the exact way of doing this depends on the time-stepping scheme 
we need a new object that takes the user-given description of the PDE and 
generates the appropriate combinations of spatial integrals and scaling.
Thus a generic problem may look as follows:
\begin{lstlisting}
  template<typename PROBLEM>
      class TimeStepProblem(Problem & P);
\end{lstlisting}
At present we have prepared the following time-stepping schemes in \dope:
\begin{itemize}
\item Forward-(or Explicit)-Euler
\item Backward-(or Implicit)-Euler
\item Crank-Nicolson scheme
\item shifted Crank-Nicolson scheme (to deal with singular initial data)
\item fractional-step-$\theta$ scheme
\end{itemize}
After choosing the time-stepping scheme, the SpaceTimeHandler  
loops over all time-steps, where in each of the 
time-steps we can reuse the functionality we have provided to solve 
stationary PDEs. 
This means for instance, if we have implemented the 
cell-integrals in the weak formulation for the stationary case, 
then we can use the exact same 
implementation again for the solution of the nonstationary version of 
the equation. However, we need to provide an additional 
term to cope with the integral for the time derivative.

@@

\subsection{Goal-oriented mesh refinement for Navier-Stokes}
In this example we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in Fig.~\ref{fig:example_ns} and the equation is discretized with the Taylor-Hood  element. We want to compute the drag force on the enclosed cylinder
\begin{equation}
J(u) = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nv _1 - pn,
\end{equation}
\todo{
\begin{equation}
J(u) = \frac 1 {20} \int_{\Gamma_{circ}} (\nu\partial_nv _1 - pn)e_1,
\end{equation}
where $e_1$ denotes the unit vector in $x$-direction. Habil 
Roland Becker S. 60 im pdf.
}
with the solution $u = (v,p)$, the normal $n$ and the viscosity $\nu = 0.001$.
\begin{figure}[hbt]
\centering
{\includegraphics[width=0.9\textwidth]{Pictures/local_grid_NS.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}
To achieve an efficient computation of the quantity $J$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}. The method delivers error indicators and by this way computational grids tailored to minimize the error in the previously given functional $J$. The price we have to pay is the additional computation of a (linear) dual problem, which is implemented as described in section~\ref{sec:opt}. In Fig.~\ref{fig:NS_local_grid} we see a resulting triangulation after eight refinement steps with roughly $320\,000$ degrees of freedom.

@@

\section{Conclusions}
In this article, we described the architecture 
of the novel software library DOpElib for solving 
PDEs and optimization problems.  The key idea of the software is to supply a
toolkit which allows for fast implementation of the problems at hand, but at
the same time gives the user full flexibility to change or alter the used
algorithms. 
This is achieved by a unified interface 
to state of the art time-stepping methods, nonlinear solvers and optimization
routines. The capabilities of DOpElib are demonstrated with the help of three examples.
